Parametricity in the Presence of GADTs

### 2021-03-02 20-gadts/GADT-story.tex
updated draft

A semanticist, on the other hand, is likely to follow~\cite{bfss90}
and expect GADTs to behave as their names suggests --- i.e., as
generalizations of ADTs. The semanticist is thus likely to expect a
GADT to be able to be used as a ADT is used --- i.e., as a shape
filled with data that can be changed without also changing the
shape. To be used in this way, a GADT must be viewed in terms of its
primitive representation. In fact, the functorial completion of a GADT
inherent in its primitive representation is exactly what is needed to
ensure that the GADT supports a $\mathsf{map}$ function, i.e., is
interpreted as a true categorical functor, and can therefore be made
an instance of Haskell's $\mathsf{Functor}$ class. It is worth noting,
however, that since the functorial completion of a GADT includes only
those data elements that are derivable via repeated applications of
$\mathsf{map}$ starting from those specified by its syntax, primitive
GADTs need not, in general, be nested types. For example, the
functorial completion of the GADT $\mathsf{G}$ defined by
\begin{equation}\label{eq:G}
\begin{array}{l}
\mathsf{data\,G\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;c :: G\,1}
\end{array}
\end{equation}
Here, $\mathsf{1}$ is the unit type, whose single element is also
denoted by $\mathsf{1}$, includes elements at any instance
$\mathsf{G\,a}$ for any type $\mathsf{a}$ that is not the empty type
$\mathsf{0}$ since there is always a function from $\mathsf{1}$ to
such an $\mathsf{a}$, but does not include any terms at instance
$\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via the
syntactic specification of $\mathsf{G}$, and is not inhabited via
functorial completion because it is not possible to define a function
from $\mathsf{1}$ into $\mathsf{0}$.

@@

Forced to choose, a programmer would likely find the idea that a GADT
contains data not specified by its syntax more than a little
disturbing. What, they might ask, should a data type contain other
than data that are constructed using its data constructors? Why should
a GADT contain ``hidden'' elements that are not specified by the
GADT's syntax and are only accessible via applications of
$\mathsf{map}$? A semanticist, on the other hand, would likely find
the primitive representation of GADTs entirely reasonable. Indeed,
they would likely find the nonfunctorial nature of a GADT's Church
encoding unnerving at best. After all, they would likely argue, the
data in a GADT shouldn't disappear just because you map a function
over it! The fact that this happens when GADTs are represented by
their Church encodings actually highlights how GADTs {\em do not}
generalize the essential, container-ish nature of ADTs at all. A
semanticist might therefore conclude that GADTs are seriously
misnamed. They might even go on to wonder: in what sense do GADTs
really deserve to be considered {\em data types} at all?

@@

The existence of parametric models that interpret GADTs as the
interpretations of their Church encodings follows from, e.g., the
existence of the parametric model of $F_\omega$ constructed
in~\cite{atk12}.  In that model, types are interpreted ``in parallel''
in $\mathit{Set}$ and $\mathit{Rel}$ in the usual way, including the
familiar ``cutting down'' of the interpretations of $\forall$-types to
just those elements that are ``parametric''~\cite{rey83,wad89} to
ensure that the Identity Extension Lemma holds. If {\color{blue} the
  set interpretation of $\mathsf{Equal}$ is the function
  $\mathit{Equal}$ and its relational interpretation is $\mathit{Eq}$}
as intended, then the parametricity property for a GADT is an
inhabitation result saying that the set interpreting any instance of
that GADT contains exactly the interpretations of the data elements
that can be formed using its data constructors and whose type is that
instance.
%Reflecting this property back into syntax gives that
In terms of syntax the parametricity property for a GADT's Church
encoding gives that the GADT is inhabited at any instance for which
data elements can be formed using its data constructors. In
particular, the parametricity property for the GADT $\mathsf{G}$
from~\eqref{eq:G} gives that $\mathsf{G\,a}$ contains a single data
element if $\mathsf{a}$ is semantically equivalent to $\mathsf{1}$ and
none otherwise:
%gives exactly the expected
%%syntactic counterpart to the
%inhabitation result for the interpretation of $\mathsf{G}$'s Church
%encoding:

\begin{example}\label{ex:CE-par}
Let $\mathsf{t}$ be a term of type $\mathsf{G}$ for the GADT
$\mathsf{G}$ defined in~\eqref{eq:G}, let $G = (G_0,G_1)$ be the
interpretation of the Church encoding of $\mathsf{G}$, let $t$ be the
interpretation of $\mathsf{t}$, and let $R :
\mathit{Rel}\,(A,B)$. Then $t\, A \in G_0\,A$ and $t\,B \in G_0\,B$
and, by the Abstraction Theorem (Theorem~3) in~\cite{atk12}, $t\,A$
and $t\,B$ must be related in $G_1\,R$. However, under the semantics
given in~\cite{atk12}, {\color{blue} which includes the aforementioned
  interpretations of $\mathsf{Equal}$}, the relational interpretation
$G_1\,R$ of $\mathsf{G}$ is {\color{blue} $\mathit{Eq}_1$ when $R$ is
  the relational interpretation $\mathit{Eq}_1$ of $\mathsf{1}$, and
  the empty relation whenever $R$ differs from
  %the relational interpretation
  $\mathit{Eq}_1$
  %of $\mathsf{1}$
  }. Reflecting back into syntax we
deduce that there can be no term in the type that is the Church
encoding of $\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically
equivalent to $\mathsf{1}$.
\end{example}

>>>>

A semanticist, on the other hand, is likely to follow~\cite{bfss90}
and expect GADTs to behave as their names suggests --- i.e., as
generalizations of ADTs. The semanticist is thus likely to expect a
GADT to be able to be used as a ADT is used --- i.e., as a shape
filled with data that can be changed without also changing the
shape. To be used in this way, a GADT must be viewed in terms of its
primitive representation. In fact, the functorial completion of a GADT
inherent in its primitive representation is exactly what is needed to
ensure that the GADT supports a $\mathsf{map}$ function, i.e., is
interpreted as a true categorical functor, and can therefore be made
an instance of Haskell's $\mathsf{Functor}$ class. It is worth noting,
however, that since the functorial completion of a GADT includes
{\color{blue} only?} those data elements that are derivable via
repeated applications of $\mathsf{map}$ starting from those specified
by its syntax, primitive GADTs need not, in general, be nested
types. For example, the functorial completion of the GADT $\mathsf{G}$
defined by
\begin{equation}\label{eq:G}
\begin{array}{l}
\mathsf{data\,G\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;c :: G\,1}
\end{array}
\end{equation}
Here, $\mathsf{1}$ is the unit type, whose single element is also
denoted by $\mathsf{1}$, includes elements at any instance
$\mathsf{G\,a}$ for any type $\mathsf{a}$ that is not the empty type
$\mathsf{0}$ since there is always a function from $\mathsf{1}$ to
such an $\mathsf{a}$, but does not include any terms at instance
$\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via the
syntactic specification of $\mathsf{G}$, and is not inhabited via
functorial completion because it is not possible to define a function
from $\mathsf{1}$ into $\mathsf{0}$.

@@

Forced to choose, a programmer would likely find the idea that a GADT
contains data not specified by its syntax more than a little
disturbing. What, they might ask, should a data type contain other
than data that are constructed using its data constructors? Why should
a GADT contain ``hidden'' elements that are not specified by the
GADT's syntax and are only accessible via applications of
$\mathsf{map}$ {\color{blue} and other elements, too?}? A semanticist,
on the other hand, would likely find the primitive representation of
GADTs entirely reasonable. Indeed, they would likely find the
nonfunctorial nature of a GADT's Church encoding unnerving at
best. After all, they would likely argue, the data in a GADT shouldn't
disappear just because you map a function over it! The fact that this
happens when GADTs are represented by their Church encodings actually
highlights how GADTs {\em do not} generalize the essential,
container-ish nature of ADTs at all. A semanticist might therefore
conclude that GADTs are seriously misnamed. They might even go on to
wonder: in what sense do GADTs really deserve to be considered {\em
  data types} at all?

@@

The existence of parametric models that interpret GADTs as the
interpretations of their Church encodings follows from, e.g., the
existence of the parametric model of $F_\omega$ constructed
in~\cite{atk12}.  In that model, types are interpreted ``in parallel''
in $\mathit{Set}$ and $\mathit{Rel}$ in the usual way, including the
familiar ``cutting down'' of the interpretations of $\forall$-types to
just those elements that are ``parametric''~\cite{rey83,wad89} to
ensure that the Identity Extension Lemma holds. If {\color{blue} the
  set interpretation of $\mathsf{Equal}$ is the function
  $\mathit{Equal}$ and its relational interpretation is $\mathit{Eq}$}
as intended, then the parametricity property for a GADT is an
inhabitation result saying that the set interpreting any instance of
that GADT contains exactly the interpretations of the data elements
that can be formed using its data constructors and whose type is that
instance. In terms of syntax the parametricity property for a GADT's
Church encoding gives that the GADT is inhabited at any instance for
which data elements can be formed using its data constructors. In
particular, the parametricity property for the GADT $\mathsf{G}$
from~\eqref{eq:G} gives that $\mathsf{G\,a}$ contains a single data
element if $\mathsf{a}$ is semantically equivalent to $\mathsf{1}$ and
none otherwise:

\begin{example}\label{ex:CE-par}
Let $\mathsf{t}$ be a term of type $\mathsf{G}$ for the GADT
$\mathsf{G}$ defined in~\eqref{eq:G}, let $G = (G_0,G_1)$ be the
interpretation of the Church encoding of $\mathsf{G}$, let $t$ be the
interpretation of $\mathsf{t}$, and let $R :
\mathit{Rel}\,(A,B)$. Then $t\, A \in G_0\,A$ and $t\,B \in G_0\,B$
and, by the Abstraction Theorem (Theorem~3) in~\cite{atk12}, $t\,A$
and $t\,B$ must be related in $G_1\,R$. However, under the semantics
given in~\cite{atk12}, {\color{blue} which includes the aforementioned
  interpretations of $\mathsf{Equal}$}, the relational interpretation
$G_1\,R$ of $\mathsf{G}$ is {\color{blue} $\mathit{Eq}_1$ when $R$ is
  the relational interpretation $\mathit{Eq}_1$ of $\mathsf{1}$, and
  the empty relation whenever $R$ differs from
  $\mathit{Eq}_1$}. Reflecting back into syntax we deduce that there
can be no term in the type that is the Church encoding of
$\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically equivalent to
$\mathsf{1}$.
\end{example}
